# Project description



## Guiding research questions



## Resources



## Your submission

Your final submission will consist of 3 separate files: the report itself (`docx` or `pdf`), one shell script `treesapp_analysis.sh`, and one R script `treesapp_analysis.R` (both script files must be in plain text format). The report should not contain any code, but should contain versions of software tools used and a high-level description of your workflow (i.e describe *what* was done and NOT *how*).

## Timeline

The following provides an outline as well as some specific milestones within the project.

```{r child = here("Rmds", "child_Rmds", "timeline.Rmd")}
```

## Reports

Reports should be formatted as per the [Instructions to Authors](https://jb.asm.org/sites/default/files/additional-assets/JB-ITA.pdf) for the [Journal of Bacteriology](https://jb.asm.org/).

Each group will submit **one** report with the sections below.

```{r child = here("Rmds", "child_Rmds", "report_structure.Rmd")}
```


