# GO term enrichment analysis

## Slides {-}

```{r, echo = FALSE, out.width = "100%", out.height = "388px"}
knitr::include_graphics("slides/tutorials/go_enrichment.pdf")
```

You can download the slides for this tutorial below.

```{r echo = FALSE}
downloadthis::download_file(
  path         = here("slides", "tutorials", "go_enrichment.pdf"),
  output_name  = "go_enrichment",
  button_label = "Download slides"
)
```

## Setting up for GO term enrichment

Create a new script or R markdown in the same project you used for the DESeq2 tutorial and install + load the required packages. Since these are Bioconductor packages, to install them you must use ```BiocManager::install```.

Load packages necessary
```{r load-packages-2, message = FALSE, warning = FALSE}
library(AnnotationDbi)
library(org.Mm.eg.db)
library(GO.db)
library(GOstats)
```

```{r, echo = FALSE}
knitr::load_cache("export-results")
```

Copy the significant results from the DESeq2 tutorial into a new file that we will be modifying.
```{r}
annotated_significant_results <- significant_res
```

The row names of your significant genes from the DEseq2 tutorial are mouse Ensembl Gene IDs. Convert the ensemble gene IDs in the rownames to Entrez IDs as a new column (and add the symbols too!)
```{r, warning = FALSE}
annotated_significant_results$symbol <- mapIds(org.Mm.eg.db,
                                 keys = rownames(annotated_significant_results),
                                 keytype = "ENSEMBL",
                                 column = "SYMBOL",
                                 multiVals = "first")

annotated_significant_results$entrez <- mapIds(org.Mm.eg.db,
                                 keys = rownames(annotated_significant_results),
                                 keytype = "ENSEMBL",
                                 column = "ENTREZID",
                                 multiVals = "first")
```

Create a non-redundant list of genes from your enriched list.
```{r, warning = FALSE}
all_genes <- annotated_significant_results %>% 
  as.data.frame() %>% 
  pull(entrez) %>% 
  unique()
```

Filter your significant genes by log2foldchange to pull out upregulated genes.
```{r, warning = FALSE}
genes_upregulated <- annotated_significant_results %>% 
  as.data.frame() %>% 
  filter(log2FoldChange > 4) %>% 
  pull(entrez) %>% 
  unique()
```

Create GO hyperGTest object from a new GOHyperGParams object that you will create with your upregulated terms and gene IDs, looking in the Biological Process ("BP") ontology.
```{r}
go_bp_upregulated <- hyperGTest(new("GOHyperGParams",
                                    geneIds = genes_upregulated,
                                    universeGeneIds = all_genes,
                                    annotation = "org.Mm.eg.db",
                                    ontology = "BP",
                                    pvalueCutoff = 0.01,
                                    conditional = FALSE,
                                    testDirection = "over"))
```

Convert your hyperGTest object into a viewable format, and print the top 10 enriched GO terms.
```{r}
go_bp_upregulated %>% summary() %>% head(10)
```

## Additional questions

Now, how could you pull out downregulated Biological Process terms in ```annotated_significant_results```? How about upregulated terms for one of the other gene ontologies?

What is the general pattern of upregulated and downregulated GO terms in the Biological Process ontology?

## Additional resources

[Gene Ontology overview](https://www.ebi.ac.uk/arrayexpress/experiments/E-MTAB-6081/samples/)
