# DESeq2: independent filtering

In this tutorial, we will explore how a large number of non-differentially expressed genes (negatives) affects the discovery of differentially expressed ones (positives). To do so, we will generate our own toy data set, 
## Slides

```{r echo=FALSE}
xaringanExtra::embed_xaringan(url = "slides/lectures/independent_filtering.html",
                              ratio = "4:3")
```

You can download the slides for this tutorial below.

```{r echo = FALSE}
downloadthis::download_file(
  path         = here("slides", "lectures", "independent_filtering.html"),
  output_name  = "7_3_deseq2_independent_filtering",
  button_label = "Download slides"
)
```

## Learning outcomes

- Explain how a large number of negatives and controlling FDR affects retention of positives.

- Produce a toy data set in R to explore the relationship among FDR-controlling procedure, negatives and positives.

- Visualize population and probability distributions in R.

- Refactor code into functions for repeated use.

## Setup

The {broom} package provides tools to format many standard R output as data frames.

```{r}
library(tidyverse)
library(broom)
```

## Generating random numbers with normal distribution 

```{r}
library(broom)
# Ctrl+shift+R
# Cmd/Ctrl_shift_P
head(rnorm(n = 1000, mean = 4, sd = 1))
```

### Creating data frames

```{r}
# Need data table for plotting
tibble(value = rnorm(n = 1000, mean = 4, sd = 1))

single_population <- tibble(population = "population_1",
                            value = rnorm(n = 1000, mean = 4, sd = 1))
```

## Visualizing data with ggplot2

```{r child = here("Rmds", "child_Rmds", "ggplot_introduction.Rmd")}
```

### Visualizing distributions 

```{r}
ggplot(single_population, aes(x = value)) +
  geom_histogram(bins = 50)

ggplot(single_population, aes(x = value)) +
  geom_density()

# Flexible with piping, adjusting values
tibble(value = rnorm(n = 1000, mean = 4, sd = 1)) %>% 
  ggplot(aes(x = value)) +
  geom_density()
```

### Setting random numbers

Random numbers in R are randomly distributed but come in a predetermined sequence. `set.seed()` allows us to select a specific set of random numbers and makes the analysis reproducible, i.e. someone else running the analysis will receive the same random numbers and thus the same results dependent on those numbers.

```{r}
set.seed(212)
tibble(value = rnorm(n = 1000, mean = 4, sd = 1)) %>% 
  ggplot(aes(x = value)) +
  geom_density()

# create 2 populations
head(c(rnorm(n = 1000, mean = 4, sd = 1),
       rnorm(n = 1000, mean = 8, sd = 1)))

# Use rep() to repeat elements
two_populations <- tibble(
  population =  rep(c("population_1", "population_2"), each = 1000),
  value      =  c(rnorm(n = 1000, mean = 4, sd = 1),
                  rnorm(n = 1000, mean = 8, sd = 1))
)

two_populations %>% 
  ggplot(aes(x = value)) +
  geom_histogram(bins = 50)
```

## Creating functions

```{r}
get_t_test_p_value_between_normal_distributions <- function() {
  sample_a <- rnorm(n = 3, mean = 4, sd = 1)
  sample_b <- rnorm(n = 3, mean = 4, sd = 1)
  result <- t.test(sample_a, sample_b)
  tidy(result)$p.value
}

p_values_for_non_differentially_expressed_genes <- tibble(
  differentially_expressed = rep(FALSE, each = 1000),
  p.value = replicate(1000, get_t_test_p_value_between_normal_distributions())
)

p_values_for_non_differentially_expressed_genes %>% 
  ggplot(aes(x = p.value)) +
  geom_histogram(boundary = 0, binwidth = 0.05)

get_t_test_p_value_between_normal_distributions <- function(n,
                                                            mean_1,
                                                            sd_1,
                                                            mean_2,
                                                            sd_2) {
  sample_a <- rnorm(n = n, mean = mean_1, sd = sd_1)
  sample_b <- rnorm(n = n, mean = mean_2, sd = sd_2)
  result <- t.test(sample_a, sample_b)
  tidy(result)$p.value
}

# Provide default values
get_t_test_p_value_between_normal_distributions <- function(n      = 3,
                                                            mean_1 = 0,
                                                            sd_1   = 1,
                                                            mean_2 = 0,
                                                            sd_2   = 1) {
  sample_a <- rnorm(n = n, mean = mean_1, sd = sd_1)
  sample_b <- rnorm(n = n, mean = mean_2, sd = sd_2)
  result <- t.test(sample_a, sample_b)
  tidy(result)$p.value
}

get_t_test_p_value_between_normal_distributions(3,
                                               mean_1 = 4, sd_1 = 1,
                                               mean_2 = 8, sd_2 = 1)

p_values_for_differentially_expressed_genes <- tibble(
  differentially_expressed = rep(TRUE, each = 1000),
  p.value = replicate(
    1000,
    get_t_test_p_value_between_normal_distributions(3,
                                                    mean_1 = 4, sd_1 = 1,
                                                    mean_2 = 8, sd_2 = 1))
)

p_values_for_differentially_expressed_genes %>% 
  ggplot(aes(x = p.value)) +
  geom_histogram(boundary = 0, binwidth = 0.05)

# combine data set
data_set <- bind_rows(p_values_for_non_differentially_expressed_genes,
                      p_values_for_differentially_expressed_genes)

data_set %>% 
  ggplot(aes(x = p.value)) +
  geom_histogram(boundary = 0, binwidth = 0.05)

# Use colour
data_set %>% 
  ggplot(aes(x = p.value, fill = differentially_expressed)) +
  geom_histogram(boundary = 0, binwidth = 0.05)

# Fix sequence of plotting
data_set %>% 
  mutate(differentially_expressed = factor(differentially_expressed,
                                           levels = c("TRUE", "FALSE"))) %>% 
  ggplot(aes(x = p.value, fill = differentially_expressed)) +
  geom_histogram(boundary = 0, binwidth = 0.05)

data_set <- data_set %>% 
  mutate(p.adj = p.adjust(p.value, method = "fdr"))

data_set %>% 
  mutate(differentially_expressed = factor(differentially_expressed,
                                           levels = c("TRUE", "FALSE"))) %>% 
  ggplot(aes(x = p.adj, fill = differentially_expressed)) +
  geom_histogram(boundary = 0, binwidth = 0.05)

simluate_p_values_for_gene_expression <- function (
    set_seed = 212,
    number_differential_genes     = 1000,
    number_non_differential_genes = 1000) {
  p_values_for_non_differentially_expressed_genes <- tibble(
    differentially_expressed = rep(FALSE, each = number_non_differential_genes),
    p.value = replicate(
      number_non_differential_genes,
      get_t_test_p_value_between_normal_distributions(3,
                                                      mean_1 = 4, sd_1 = 1,
                                                      mean_2 = 4, sd_2 = 1))
  )
  p_values_for_differentially_expressed_genes <- tibble(
    differentially_expressed = rep(TRUE, each = number_differential_genes),
    p.value = replicate(
      number_differential_genes,
      get_t_test_p_value_between_normal_distributions(3,
                                                      mean_1 = 4, sd_1 = 1,
                                                      mean_2 = 8, sd_2 = 1))
  )
  bind_rows(p_values_for_non_differentially_expressed_genes,
            p_values_for_differentially_expressed_genes) %>% 
    mutate(p.adj = p.adjust(p.value, method = "fdr"))
  
}

data_set <- simluate_p_values_for_gene_expression()
data_set %>%
  mutate(differentially_expressed = factor(differentially_expressed,
                                           levels = c("TRUE", "FALSE"))) %>% 
  ggplot(aes(x = p.adj, fill = differentially_expressed)) +
  geom_histogram(boundary = 0, binwidth = 0.05)


simluate_p_values_for_gene_expression(number_differential_genes = 1000,
                                      number_non_differential_genes = 5000) %>% 
  mutate(differentially_expressed = factor(differentially_expressed,
                                           levels = c("TRUE", "FALSE"))) %>% 
  ggplot(aes(x = p.adj, fill = differentially_expressed)) +
  geom_histogram(boundary = 0, binwidth = 0.05)

# At one point do you not have any more True positives
```

## Additional resources

### Independent filtering:

- [StatQuest: Filtering genes with low read counts](https://statquest.org/statquest-filtering-genes-with-low-read-counts/)

### ggplot2

```{r child = here("Rmds", "child_Rmds", "ggplot_resources.Rmd")}
```
