# {DESeq2}: model fitting

```{r include = FALSE}
library(here)
```

## Slides

```{r echo=FALSE}
xaringanExtra::embed_xaringan(url = "slides/lectures/model_fitting.html",
                              ratio = "4:3")
```

You can download the slides for this tutorial below.

```{r echo = FALSE}
downloadthis::download_file(
  path         = here("slides", "lectures", "model_fitting.html"),
  output_name  = "7_4_deseq2_model_fitting",
  button_label = "Download slides"
)
```

## Learning outcomes

- Describe properties of read count data and how DESeq models them.

- Use R Markdown to create reproducible documents.

## Setup

### R packages

Install the R Markdown package.

```{r eval = FALSE}
install.packages("rmarkdown")
```

## What is R Markdown?

R Markdown allows you to create reproducible documents containing code. It both includes the information on content and formatting information "markup". R Markdown offers a simple way to markup your text using a simplified markup language called "Markdown".

### Formatting for emphasis

You get `*`*emphasis*`*` with single asterisk, and `**`**strong emphasis**`**` with two asterisks.

### Superscript and subsript

superscript`^`^2^`^` subscript`~`~2~`~`

### Linbreaks and paragraphs

A regular new line won't render as a new line in the formatted document. Ending a line with two spaces creates a line break  
like so. Leaving a blank line, creates a new paragraph.

This is a new paragraph.

## Themes

[Check out different Bootstrap themes](https://bootswatch.com/3/). You can use them in the YAML header with `theme:` followed by default, cerulean, journal, flatly, darkly, readable, spacelab, united, cosmo, lumen, paper, sandstone, simplex, or yeti.

## Additional resources

- [Interactive tutorial for learning Markdown](https://commonmark.org/help/tutorial/)

- [Publishing with Netlify Drop](https://app.netlify.com/drop)