# Working with SAMtools and BCFtools Part 2

## Slides {-}

```{r, echo = FALSE, out.width = "100%", out.height = "388px"}
knitr::include_graphics("slides/tutorials/samtools_bcftools_part_2.pdf")
```

You can download the slides for this tutorial below.

```{r echo = FALSE}
downloadthis::download_file(
  path         = here("slides", "tutorials", "samtools_bcftools_part_2.pdf"),
  output_name  = "bwa_samtools_and_bcftools_part_2",
  button_label = "Download slides"
)
```

## Main Exercise {-}

1. Create a new alignment using ```bwa aln```
2. Do a quick comparison using the SN metrics from ```samtools stats``` of the ```bwa aln``` BAM and the previous ```bwa mem``` BAM. What is one metric that changed the most and why might this be?
3. Filter the ```bwa aln```alignment in order to only keep reads with a mapping quality above a threshold of your choice.
  + Do a quick side-by-side comparison of coverage of your new and old ```bwa aln``` alignments using ```bwa stats```. How does introducing a quality threshold lead to a difference here? Can you explain the pros and cons of filtering by mapping quality in terms of [sensitivity and specificity](https://en.wikipedia.org/wiki/Sensitivity_and_specificity)?
  + Choosing an arbitrary area along *Bordetella*'s single contig, what is one difference you can spot between your new and old ```bwa aln``` alignments besides read depth? (Hint: take a look at bit flags and positions, and [consult IGV's coloring scheme](http://software.broadinstitute.org/software/igv/interpreting_insert_size))

## Additional resources {-}

[BWA manual pages](http://bio-bwa.sourceforge.net/bwa.shtml)

[SAMtools manual pages](http://www.htslib.org/doc/samtools.html)

[BCFtools manual pages](http://www.htslib.org/doc/bcftools.html)

[Introduction to the BAM file format](https://github.com/davetang/learning_bam_file)

[Introduction to the VCF file format](https://github.com/davetang/learning_vcf_file)

[IGV download](https://software.broadinstitute.org/software/igv/download)

[IGV user guide](http://software.broadinstitute.org/software/igv/UserGuide)
