## Data structures

The data types in R are the elements to build more complex objects called *data structures*. The most basic structures are *vector*, *lists*, *matrix* and *data frame*.

### Vector

We can use the `c()` function to create a vector. All vector elements have to be of the same data type (a so-called *atomic* vector).

```{r}
vector_example <- c("This", "vector", "has", "5", "elements!")
vector_example
```

::: rmdwarning
**Coercion**

Since all vector elements have to be of the same type, what happens if we use `c()` with dissimilar ones? To test R's behaviour, create several vectors of all possible pairwise combinations of `character`, `numeric` (use any integer), and `logical`, then determine the type of the object using `typeof(<your vector>)`. In addition, test the combination of (a) `logical` with either the numbers `0` or `1` or (b) `logical` and an integer larger than 1 or smaller than -1.

The behaviour you will observe is called *coercion*. Why do you think this particular set of behaviours were selected for the design of R?
:::

You can also assign each element a name using `<name> = <value>`

```{r}
vector_example <- c(first = "Hi", second = "there")
vector_example
```

### List

In R, a list is a special kind of vector whose elements can be any combination of objects (i.e. they also do not have to be the same).

```{r}
list_example <- list(c("gene_a", "gene_b"),
                     c("untreated", "control"))
```

You can also assign a name to elements of a list.

```{r}
list_example <- list(gene_names = c("gene_a", "gene_b"),
                     treatments = c("untreated", "control"))
```

### Matrix

The data is provided as a vector to `matrix()` and filled either column- or row-wise into the matrix. Rows and columns can also be assigned names.

```{r}
matrix_example <- matrix(
  data = c(
    10, 20,
    3,  6
  ),
  byrow = TRUE,
  ncol = 2,
  dimnames = list(
    c("gene_a", "gene_b"),
    c("control", "treated")
  )
)

matrix_example
```

### Data frame

A data frame is another data structure that contains elements in columns and rows and it is a list of vectors, with each column being one vector of a single data type but the data type may differ among vectors. Data frames are a big part of working with data in R. Data is always provided by column `<column name> = <data>`

```{r}
data_frame_example <- data.frame(
  gene_name = c("gene_a", "gene_b"),
  control = c(10, 3),
  treated = c(20, 3)
)

data_frame_example
```

Although the data frame and matrix conain the same information, they are not the same *class* of object. You can determine the class with `class()`.

```{r}
class(matrix_example)
class(data_frame_example)
```

### Subsetting

We often want to use only part of the data and have to subset them accordingly. In base R, subsetting is done with brackets `[ ]`. For example, a vector can be subset by the index or the name of the position.

```{r}
vector_example[1]
vector_example["first"]
```

Matrices and data frames can also be subset that way by first specifying rows then columns `[<rows>, <columns>]`.

```{r}
matrix_example[,  "control"]
matrix_example[1, "control"]
```

We can extract a single column (vector) from a data frame using the `$` operator.

```{r}
data_frame_example[, "control"]
data_frame_example$control
```
