## Data types

When working with data in R, you might have noticed that not all data are the same, and these different forms of data are called *data types*. As part of this tutorial, you will encounter three basic data types:

| Data type | Description                                                  |
|-----------|--------------------------------------------------------------|
| character | Any type of text. Must be enclosed in quotes `"<your text>"` |
| numeric   | Decimal numbers                                              |
| logical   | Can only take `TRUE` or `FALSE`                              |

Let's look at a few examples of different data types. Below is a simple example of the type *character*. Character types are always enclosed by quotes.

```{r}
salutation <- "Hello, World!"
salutation
```

Here an example of a *numeric*:

```{r}
some_number <- 6
some_number
```

It is important to use the correct data type for a function or other operations in R. For example, the operator `+` only works with numeric data

```{r}
6 + 2
```

but not character data

```{r error = TRUE}
"6" + "2"
```

*Logicals* can take either `TRUE` or `FALSE`. You might have already encountered them as values for arguments in functions.

```{r}
microbiology_is_awesome <- TRUE
microbiology_is_awesome
```
