---
title: "Tutorial 2 --- Bash & scripting"
subtitle: "MICB 405 --- Bioinformatics 2021W1"
author: "Axel Hauduc"
institute: "University of British Columbia"
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

```{r setup, echo=FALSE}
xaringanExtra::use_share_again()
xaringanExtra::style_share_again(
  share_buttons = c("twitter", "linkedin", "pocket")
)
```

## The Unix Philosophy

1. Make each program do one thing well. To do a new job, build a new program rather than complicate old programs.

1. The output of any program should easily become the input for another.

---

## Utilities review

`cd` change directory

`ls` list files (`-l` long)

`pwd` print working directory

`cp` copy

`rm` remove (`-r` recursive)

`mv` move

`cat` concatenate

`echo` print content you provided as argument

---

## More utilities...

`grep` global regular expression print

`less` read text files

`head`/`tail` print first/last

`wc` wordcount `-l` lines, `-w` word, `-m` character

`sort` sort lines

`uniq` return unique lines

`chmod` change mode a.k.a. permissions

`mkdir` make directory

---

## `awk` command

Less Unix-y than previous commands

More of a multi-tool for dealing with data files

Not short for anything – does file processing based on lines & columns

Follows the format:

`awk '<filter> { <action }' <file>`
  
e.g.
  
`awk 'NR >= 2 { print $0 }' mtcars.tsv`

---

## Special characters

`>` redirect

`>>` append

`|` pipe (not `1`, `l`, or `I`)

`;` separate command

`&&` AND

`||` OR

`*` wildcard

---

## In-terminal text editing

`nano` = easiest to use & sufficient for most purposes

Example: `nano <filename>`

command creates a new file, or edits an existing one of that name

<kbd>ctrl</kbd> + <kbd>o</kbd> overwrite (i.e. save)

<kbd>ctrl</kbd> + <kbd>x</kbd> exit

---

## Scripting

Helps repeat repetitive tasks

Scripts the form of a text file that can be "executed" when you need them

---

## Script headers

All script files need a header to indicate how the script file should be run

First line specifies the program that will interpret the rest of the script

```{bash eval = FALSE}
#!/bin/bash
```

- Program should interpret following text using Bash

- Other programs = other headers, if run by calling the script directly

---

## Script variables

We can create a variable and assign it a value with

```{bash eval = FALSE}
results_dir="results/"
```

Note that spaces matter when setting Bash variables. Do **NOT** use spaces around the equal sign `=`.

---

## Variables


To access a variable's value, we use a dollar sign in front of the variable's name.

Suppose we want to create a directory for a sample's alignment data, called `<sample>_aln/`, where `<sample>` is replaced by the samples name.

```{bash eval = FALSE}
sample="Individual_2A"
mkdir "${sample}_data/"
```

This will create a directory with the name `Individual_2A_data`.
Curly braces `{ }` indicate where variable name starts and ends.

---

## Command line arguments

```{bash eval = FALSE}
grep -c "<string>" <file>

bash <script> <path/to/file>
```
Arguments added after calling a script are assigned to the default variables `$1`, `$2`, `$3`, and so forth within the script.

---

## `for` loops

In bioinformatics, most of our data is split across multiple files.

Many processing pipelines need a way to apply the same workflow on each file, taking care to keep track of sample names.

Looping over files with Bash’s `for` loop = simplest way to accomplish this

Three essential parts to creating a pipeline to process a set of files:

1. Selecting which files to apply the commands to
1. Looping over the data and applying the commands
1. Keeping track of the names of any output files created

---

## `for` loops template

```{bash eval = FALSE}
#!/bin/bash
for a_file in </path/to/dir/>*
do
  <action on> $a_file
  <another action on> $a_file
done
```

---

## `for` looping through directories

Creating for loops that loop through an entire directory

```{bash eval = FALSE}
#!/bin/bash

for foo in /home/ahauduc_mb20/test_directory/*
do
  head $foo > ${foo}.head.and.tail.txt
  tail $foo >> ${foo}.head.and.tail.txt
done
```

---

## `if` statements

Use if you want to perform commands on a subset of files, or only if an action meets certain conditions.

The basic syntax is:

```{bash eval = FALSE}
if <condition is true>
then
  <DO THIS>
else
  <DO THAT>
fi
```

---

## `if` statements

```{bash eval = FALSE}
#!/bin/bash

if cat $1
then
  echo "The file exists!"
else
  echo "The file doesn't exist!"
fi
```

---

## Return codes

`0` Command executed successfully

`1`+ Command did not execute successfully

There can by many different error types denoted by specific numbers

---

## Return codes are invisible

```{bash eval = FALSE}
#!/bin/bash

if cat $1
then
  echo "The file exists!"
else
  echo "The file doesn't exist!"
fi
```

---

## `test` statements: `[[ <condition> ]]`

Like other programs, `test` exits with either `0` or `1`.

Test statements can be included at the beginning of the `if` program to make producing the right return code easier

`test` supports numerous helpful comparisons you might need

---

## `test` summary comparisons

```{r echo = FALSE, message = FALSE}
readr::read_tsv("../../data/test_summary.tsv") |>
  dplyr::slice(1:9) |>
  knitr::kable(format = 'html', table.attr = "style='width:100%;'") |>
  kableExtra::column_spec(1, width = "25%")
```

---

## `test` summary for files/directories

```{r echo = FALSE, message = FALSE}
readr::read_tsv("../../data/test_summary.tsv") |>
  dplyr::slice(10:15) |>
  knitr::kable(format = 'html', table.attr = "style='width:100%;'") |>
  kableExtra::column_spec(1, width = "25%")
```

---

## `if` + `test` statements

Combining `test` with `if` statements is simple:

```{bash eval = FALSE}
#!/bin/bash
if [[ <condition> ]]
then
  <this action>
else
  <that action>
fi
```

Note the spaces around and within the brackets `[[ ]]`: these are required.

---

class: middle

## Putting it all together...

---

## `for` looping through directories

```{bash eval = FALSE}
#!/bin/bash

for foo in /home/axel/Documents/data/*
do
  if [[ -f ${foo} ]]
  then
    head ${foo} > ${foo}.head.and.tail.txt
    tail ${foo} >> ${foo}.head.and.tail.txt
  else
    echo “${foo} is not a file to summarize!”
  fi
done
```
